---
title: "Exploratory analysis of the Albury time series data"
author: Anna Quaglieri
output: html_document
---


# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      prompt = TRUE,
                      warning = FALSE, fig.path = "./figures/")

library(tidyverse)
library(lubridate)
library(tsibble)
library(fable)
library(feasts)
library(cowplot)
library(here)
library(hexbin)
library(viridis)
```

# Water level data

The data below refer to the historical data set of water at Murray basin at Albury back from 1885 (lat -36.0981, long 146.9065). The dataset is available through the Amphora website at https://beta.amphoradata.com/Amphorae/Detail?id=5602eb66-fb21-4350-9f68-0e5cbb925f62. 


```{r create-data-tsibble,include=FALSE}
Albury_historical <- read_csv(file.path(here(),"data/Albury_historical.csv"), col_names = FALSE, skip = 4,
                              col_types = cols(X3 = col_double(), X4 = col_double()))
summary(Albury_historical)

# Convert first column to readable dates
albury_time <- Albury_historical %>%
  separate(X1, into = c("Time","Date"),sep=" ",remove = FALSE) %>%
  mutate(Date = as.Date(Date,"%d/%m/%Y"),
         Time = hms(Time)) 

albury_time <- albury_time %>%
  dplyr::select(Date,X2,X3,X4) %>%
  dplyr::rename(wlevel = X2,
                wsal = X3,
                wtemp = X4) %>%
  mutate(origin = "albury")

twater <- albury_time %>% 
  as_tsibble(index = Date)

twater <- twater %>%
  mutate(interval = case_when(Date < "1900-01-01" ~ "Before 1900",
                              Date < "1940-01-01" & Date >= "1900-01-01"~ "1900-1940",
                              Date >= "1940-01-01" & Date < "1980-01-01" ~ "1940-1980",
                               Date >= "1980-01-01" ~ "1980-today",)) %>%
  mutate(interval = factor(interval,levels=c("Before 1900","1900-1940","1940-1980","1980-today")))
```


# Time series of water level/salinity/temperature data

```{r all-level-ts}
p1=twater %>%
  feasts::autoplot(wlevel) + ggtitle("Water level at Albury along the Murray river") + theme_classic()+ labs(x = "Years",y="Level (m)")

p2=twater %>%
  feasts::autoplot(wtemp) + theme_classic()+ labs(x = "Years",y="Temperature (Deg C)")

p3=twater %>%
  feasts::autoplot(wsal) +  theme_classic() + labs(x = "Years",y="Salinity (uS/cm)")

plot_grid(p1,p2,p3,nrow=3)
```


```{r hist-wlevel, fig.height=3,fig.width=10}
twater %>%
  feasts::autoplot(wlevel) + 
   autolayer(filter_index(twater, "2001-01-01" ~ "2007-12-31"), color = "#de2d26") +
  ggtitle("Water level at Albury along the Murray river. Drought years 2001-2007 highlighted in red.") + 
  theme_bw()+ labs(x = "Years",y="Level (m)")
```


# Zoomed time series of water salinity/temperature data

```{r ts-after-2000}
p3=twater %>%
  filter(year(Date) > 2000) %>%
  autoplot(wlevel) + ggtitle("Water level after year 2000")  + theme_bw()+
   autolayer(filter_index(twater, "2015-01-01" ~ "2016-01-01"), color = "#de2d26") +
  labs(x = "Years",y="Level (m)")

p1=twater %>%
  filter(year(Date) > 2000) %>%
  autoplot(wtemp) + ggtitle("Water temperature after year 2000")  + theme_bw()+
    autolayer(filter_index(twater[,c("Date","wtemp")], "2015-01-01" ~ "2016-01-01"), color = "#de2d26") +
  labs(x = "Years",y="Temperature (Deg C)")

p2=twater %>%
  filter(year(Date) > 2000) %>%
  autoplot(wsal) + ggtitle("Water salinity after year 2000")  + theme_bw() +
    autolayer(filter_index(twater[,c("Date","wsal")], "2015-01-01" ~ "2016-01-01"), color = "#de2d26") +
  labs(x = "Years",y="Salinity (uS/cm)")

plot_grid(p3,p1,p2,nrow=3)
```

# Time series of water level data by year

```{r water-level-year-interval}
twater %>%
  gg_season(wlevel, colour = "black",alpha=0.5) + ggtitle("Water level time series by intervals") + 
  theme_bw() +
  facet_wrap(~interval) + theme(legend.position = "none")+ labs(x = "Months",y="Level (m)")
```


```{r water-sal-2009}
twater %>%
  filter(year(Date) > 2009) %>%
   mutate(year = year(Date)) %>%
  gg_season(wsal,colour="black") + ggtitle("Water Salinity, between 2010-2019")  + theme_bw()+
  labs(x = "Years",y="Salinity (uS/cm)")+ theme(legend.position = "none") +
  facet_wrap(~year)

```


```{r water-temp-after2000}
twater %>%
  filter(year(Date) > 2000) %>%
   mutate(year = year(Date)) %>%
  gg_season(wtemp,alpha=0.7) + 
  ggtitle("Water temperature, between 2001-2019")  + theme_bw()+
  labs(x = "Years",y="Temperature (Deg C)") + theme(legend.position = "bottom")
```


# Correlations between water level and temperature/salinity

```{r watertemp-waterlev}
# Water levels vs water temperature
twater %>%
    mutate(year = year(Date),
          month = month(Date),
          day = day(Date)) %>%
  filter(year >= 2000) %>%
  mutate(Year = factor(year(Date))) %>%
  ggplot(aes(x = wlevel, y = wtemp,colour=Year)) + 
  geom_point(alpha=0.3,colour="grey") +
  geom_smooth(se=FALSE)+
  theme_bw() +
  ggtitle("Water level vs water temperature (>2000)") +
  labs(x = "Water level (m)", y = "Water temperature (Deg C)") +
  theme(legend.position = "bottom")+
  scale_color_viridis_d()
```


```{r watersal-waterlev}
# Water levels vs water salinity
twater %>%
  filter(year(Date) > 2009) %>%
  mutate(Year = factor(year(Date))) %>%
  ggplot(aes(x = wlevel, y = wsal,colour=Year)) + 
  geom_point(alpha=0.3) +
  geom_smooth(se=FALSE) +
  theme_bw()+
  labs(x = "Water level (m)", y = "Water salinity (uS/cm)") +
   ggtitle("Water level vs water salinity (>2010)") 

```


```{r water-lev-salt-month}
# Water levels vs water salinity
twater %>%
  filter(year(Date) > 2000) %>%
  mutate(Year = factor(year(Date)),
         Month = factor(month(Date))) %>%
  ggplot(aes(x = wlevel, y = wsal,colour=Month)) + 
  geom_point(alpha=0.5) +
  theme_bw()+
  labs(x = "Water level (m)", y = "Water salinity (uS/cm)") +
   ggtitle("Water level vs water salinity (>2010)")
```

# Scatterplot without stratifying by year/month

```{r density-sal-water}
# Water levels vs water salinity
twater %>%
  filter(year(Date) > 2000) %>%
  mutate(Year = factor(year(Date))) %>%
  ggplot(aes(x = wlevel, y = wsal)) + 
   geom_hex(bins = 70) +
  scale_fill_continuous(type = "viridis") +
  theme_bw()+
  labs(x = "Water level (m)", y = "Water salinity (uS/cm)") +
   ggtitle("Water level vs water salinity (>2010)")

```


```{r density-temp-water}
# Water levels vs water salinity
twater %>%
  filter(year(Date) > 2000) %>%
  mutate(Year = factor(year(Date))) %>%
  ggplot(aes(x = wlevel, y = wtemp)) + 
   geom_hex(bins = 70) +
  scale_fill_continuous(type = "viridis") +
  theme_bw()+
  labs(x = "Water level (m)", y = "Water temperature (Deg C)") +
   ggtitle("Water level vs water temperature (>2010)")

```


```{r density-sal-water-nodens}
# Water levels vs water salinity
plot_sal <- twater %>%
  filter(year(Date) > 2000) %>%
  mutate(Year = factor(year(Date))) 

plot_sal %>%
  ggplot(aes(x = wlevel, y = log2(wsal))) + 
  geom_point(alpha=0.3) +
  geom_smooth(se=FALSE) +
  geom_point(data = subset(plot_sal, Year %in% "2015"),colour="#de2d26",alpha=0.3) +
  theme_bw()+
  labs(x = "Water level (m)", y = "Water salinity (uS/cm)") +
   ggtitle("Water level vs water salinity (>2010).\nYear 2015 highlighted in red")
```


```{r}
# Water levels vs water salinity
twater %>%
  filter(year(Date) > 2000) %>%
  mutate(Year = factor(year(Date)),
         Month = factor(month(Date))) %>%
  ggplot(aes(x = wlevel, y = wsal)) + 
   geom_hex(bins=20) +
  scale_fill_continuous(type = "viridis") +
  theme_bw()+
  facet_wrap(~Month) +
  labs(x = "Water level (m)", y = "Water salinity (uS/cm)") +
   ggtitle("Water level vs water salinity (>2010)")
```


# Min/max water temperature

```{r}
min_max <- twater %>%
  as.data.frame() %>%
  mutate(Year = year(Date)) %>%
  filter(Year > 2000) %>%
  group_by(Year) %>%
  summarise(min_temp = min(wtemp,na.rm=TRUE),
            max_temp = max(wtemp, na.rm=TRUE)) %>%
  gather(key = summary, value = wtemp, min_temp,max_temp)
```


```{r min-max-ts}
ggplot(min_max, aes(x = Year, y = wtemp,colour=summary)) + geom_point() +
  geom_line() + theme_bw()
```


# Lag plots

```{r}
p1.data <- twater %>% 
  filter(year(Date) > 2017) %>%
  mutate(lag30.wlevel = c(diff(wlevel,lag=30),rep(NA,30)),
         lag30.wtemp = c(diff(wtemp,lag = 30),rep(NA,30))) %>%
  gather(key = measure, value = level,lag30.wlevel:lag30.wtemp)

p1=twater %>%
   filter(year(Date) > 2018) %>%
  autoplot(wlevel)
p2=twater %>%
   filter(year(Date) > 2018) %>%
  autoplot(wtemp)
plot_grid(p1,p2,nrow=2)
```

